From physical conditions to Hamiltonian and Lagrangian mechanics

### 2021-02-07 Articles/2020-ClassicalMechanicsPhil/Source/classical-phil-article.tex
Update classical-phil-article.tex
edits to all sections but 5.1 (coordinate-dependence) and Curiel section [before merging with GC edits from Monday]

Philosophers have recently shown that we can interpret Lagrangian and Hamiltonian mechanics as theoretically equivalent in a precise mathematical sense. Within a subclass of classical mechanical systems known as the \textit{hyper-regular domain}, the two formalisms are categorically equivalent \parencites[]{Teh}{Barrett2}. Categorical equivalence provides a natural interpretive criterion for theoretical equivalence because it yields mutual inter-translatability between theory formulations. At first glance, this equivalence motivates the view that there are no physically significant differences between these two formalisms, at least within the hyper-regular domain. Yet, it allows for physical and conceptual differences to arise outside the hyper-regular domain. 
%Once identified, these differences can even affect how we interpret hyper-regular systems.

Within this wider domain, \textcites[]{North} has argued for privileging Hamiltonian mechanics, on the grounds that it requires less mathematical structure to formulate. In contrast, \textcites[]{Curiel} recommends privileging Lagrangian mechanics, arguing that it isolates physically-significant kinematical constraints that all classical systems must satisfy. Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding. Whereas North's argument relies on a problematic criterion for comparing structure---based on the size of symmetry groups \parencites[]{Swanson}---our argument avoids this pitfall. We show that a system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics, and this difference has interpretive consequences even within the hyper-regular domain. Furthermore, the construction at the heart of this argument provides an interpretation of classical mechanical systems in terms of their invariant phase space density, along with a novel interpretation of the principle of stationary action within the Hamiltonian formalism.

Understanding the relationship between Hamiltonian and Lagrangian mechanics depends partly on how one defines ``classical systems." Different definitions can result in different interpretive relationships. For instance, restricting classical systems to the hyper-regular domain precludes certain interpretive considerations. To avoid verbal disputes, it is essential to clarify what falls within the scope of ``classical systems." One way to delimit classical systems is to specify a set of equations that they satisfy---such as Hamilton's equations or the Euler--Lagrange equations. We will instead follow a different approach by delimiting classical systems based on the physical conditions that they satisfy. Section~\ref{infinitesimal} introduces the condition of \textit{infinitesimal reducibility}, meaning that the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (we cannot specify the whole state by specifying the states of its subsystems). In the widest sense, a classical system is any infinitesimally reducible system.
%a reducible system comprises subsystems whose states jointly determine the state of the whole system.

Section~\ref{deterministic} introduces the joint condition of deterministic and reversible evolution. We will show that a classical system which satisfies this condition is Hamiltonian: its phase space is a symplectic manifold that evolves according to Hamiltonian evolution  (i.e. according to a symplectomorphism). Following a deterministic and reversible evolution is equivalent to being a conservative system, i.e. one that conserves energy. Hence, our discussion of Hamiltonian mechanics necessarily excludes dissipative systems, i.e. ones that do not conserve energy. Likewise, our derivation of Lagrangian mechanics in Section~\ref{motion} will also take place within the context of conservative systems. Consequently, we will not consider systems that satisfy inhomogenous Euler--Lagrange equations. We, of course, do not deny the sense in which the Lagrangian formulation applies to dissipative systems, but this is not what we will mean by ``Lagrangian mechanics" \parencites[]{Smith}{Curiel}.\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian.

@@

We begin in Section~\ref{privileging} with an informal overview of our argument for privileging Hamiltonian mechanics. We introduce the key physical conditions that distinguish Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. We then show (Section~\ref{motion}) that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. In contrast, we show that in configuration space, the invariance of the density is coordinate-dependent (arising from the fact that the Jacobian equals unity in Cartesian coordinates but not in general). Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics.
% In configuration space, it is in some sense an accident that the density is invariant

@@

We now outline the three main steps of our argument, summarized in Figure~\ref{diagram}. The left column lists three basic physical conditions: infinitesimal reducibility, determinism/reversibility, and motion equivalence. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal mathematical framework depicted in the right column: including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby gain a straightforward interpretation of the physical significance of these formalisms.
%Of the various notions of fundamentality, there is at least the following non-mysterious definition:

@@

First, a system is \textit{infinitesimally reducible} provided that the states of its constituent, infinitesimal parts completely specify the overall state of the system. Colloquially, we can say that the whole state is equal to the sum of its parts, where we take these infinitesimal parts to be classical particles. When this condition is satisfied, we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that to define this invariant distribution, we require a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility. [To do: Clarify the scope of our treatment: is it restricted to $N$-particle classical mechanics? Or does it also handle rigid bodies and classical continual mechanics? Cite Mark Wilson 2013 on these issues]. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

Secondly, a system has \textit{deterministic and reversible evolution} provided that the present state of the system uniquely determines both its future and past states. This condition entails that the invariant distribution $\rho$ is in fact a conserved phase space density. We can thus model such systems using Hamiltonian mechanics, defined on a phase space with a symplectomorphism. These two conditions thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution. Section~\ref{deterministic} justifies this correspondence. [To do: on the topic of failures of determinism, cite Earman  \textit{a primer on determinism}, Baez singularities paper, or Norton's dome paper].

Finally, some classical systems satisfy a further condition, which we call \textit{motion equivalence}. Motion equivalence holds when the system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics.

Importantly, some systems fail to satisfy the condition of motion equivalence. It is not always true that we can determine the evolution of a system in phase space from its trajectory in physical space. This is because knowing a system's position and velocity is not in general sufficient for knowing its momentum. For instance, knowledge of the physical trajectory of a photon, when treated as a particle, is insufficient for knowledge of its phase space evolution, since we cannot determine the momentum of a photon from only its trajectory through space. In general, different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
[To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises only when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

@@

% Overall summary: A classical system is an infinitesimally reducible system. The state space of these systems can be modeled by a symplectic manifold (the physical assumption of infinitesimal reducibility corresponds to the mathematical structure of symplectic manifolds). Classical Hamiltonian systems satisfy the additional conditions of having deterministic and reversible evolutions. These two physical assumptions entail the existence of a symplectic morphism on the state space manifold, making it a contact manifold. Classical Lagrangian systems satisfy a further assumption: keeping track of the trajectories in physical space is equivalent to tracking the evolution of states in phase space.

\subsection{(Non-metaphysical) fundamentality and structure}
\label{fundamentality}

% % To do: slip in a citation to Dorr/Hawthorne 2014, to note the independent metaphysical challenges that the notion of naturalness faces, and likewise for a metaphysical notion of fundamentality.

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. On our usage, differences in fundamentality amount to differences in generality.\footnote{Of course, we allow for other non-metaphysical uses of ``fundamental," tracking differences in epistemic or conceptual primacy. For instance, we can think of the canonical position and momentum observables as ``epistemically fundamental" because they suffice to completely specify a classical state, fixing its values for all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." Section~\ref{fundamental}, discusses further senses in which the Hamiltonian formalism is more fundamental than Lagrangian mechanics.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires that fewer physical conditions are satisfied. Compared with North's \parencites*[]{North} account of structure, our definition has two distinct advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

Methodologically (but not metaphysically), our approach has many parallels to North's. We agree with North that it is fruitful to identify both invariants and---more generally---``the mathematical structure needed to formulate the theory in an invariant, frame-independent way" \parencites*[65]{North}. Additionally, we agree that it is worthwhile to identify what particular mathematical structure is required by a formulation \parencites[78]{North}. Indeed, we undertake precisely this sort of investigation in Section~\ref{derivation}. However, North advocates a metaphysical account of structure, about which we remain agnostic. According to North, ``structure comprises the objective, fundamental, intrinsic features, the ones that remain the same regardless of arbitrary or conventional choices and description" \parencites*[66]{North}. These notions of fundamentality and intrinsicality correspond to differences in joint-carving, with more fundamental or intrinsic features being more joint-carving, or in Lewis's \parencites*[]{Lewis1983} terminology, more natural.

Owing to our epistemic inabilities to access these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant distributions (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus structure. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. As we will see, precisely this scenario occurs in the context of Lagrangian mechanics. Hence, whereas North views Lagrangian mechanics as containing ``excess, superfluous structure" \parencites*[75]{North}, we take this additional structure as unnecessary but still physically significant. It corresponds to the physical assumption of motion equivalence, which applies to many classical systems. 
%This characterizes the relationship between Hamiltonian and Lagrangian formalizations of a classical system. 
% classical mechanics also deploys physically significant representations that are not intrinsic to mechanical systems [to do: mention some examples, e.g. in a footnote. The Lagrangian? Cartesian coordinates?].

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool.} Whether Hamiltonian phase space structure is metaphysically fundamental does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the at least three different mathematical ways of formalizing Hamiltonian mechanics [mention these three formalisms and provide references]. Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, unique necessary mathematical structure for mechanics, but rather a family of related structures. There might not be a single canonical minimal structure for formalizing the relevant physical models and physical assumptions. Furthermore, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

>>>>

The intellectual richness of classical mechanics manifests itself in an abundance of formulations: with Newtonian, Hamiltonian, Lagrangian, and Hamilton--Jacobi being the most prominent.  Butterfield \parencites*[]{Butterfield} has aptly referred to these formulations as different \textit{frameworks} for treating classical systems. Recent work in philosophy of science has focused largely on how these formulations relate to one another, with the aim of arriving at a deeper understanding of classical mechanics. By considering the physical and conceptual differences between these formulations, we can better distinguish physical features from artifacts of our representations.

In the context of comparing Hamiltonian with Lagrangian mechanics, three interpretive stances have emerged. \textcites[]{North} has argued for privileging Hamiltonian mechanics, on the grounds that it requires less mathematical structure to formulate. In contrast, \textcites[]{Curiel} recommends privileging Lagrangian mechanics, arguing that it isolates physically-significant kinematical constraints that all classical systems must satisfy. Finally, we can interpret Lagrangian and Hamiltonian mechanics as theoretically equivalent in a precise mathematical sense. Within a subclass of classical mechanical systems known as the \textit{hyper-regular domain}, the two formalisms are categorically equivalent \parencites[]{Teh}{Barrett2}. Categorical equivalence provides a natural interpretive criterion for theoretical equivalence because it yields mutual inter-translatability between theory formulations. This equivalence motivates the view that there are no physically significant differences between these two formalisms, at least within the hyper-regular domain. 
% It is interesting to think how the other formulations fit in, e.g. Hamilton--Jacobi. Is there a categorical equivalence result for this formulation in some domain?

Here, we will argue that there are indeed physical grounds for privileging Hamiltonian mechanics over Lagrangian mechanics, their categorical equivalence notwithstanding.\footnote{Barrett shows that in the hyper-regular domain, Lagrangian and Hamiltonian mechanics are categorically equivalent provided that we define categories of models over the tangent and cotangent bundle formulations, respectively \parencites*[1181-82]{Barrett2}. However, he also shows that if we consider a more general formulation of Hamiltonian mechanics on a symplectic manifold (which does not require a cotangent bundle), then the two formulations are \textit{not} categorically equivalent \parencites*[1182-83]{Barrett2}. Our construction of Hamiltonian mechanics in Section~\ref{derivation} matches this more general symplectic manifold formulation, thereby avoiding trivialization worries from categorical equivalence.} Whereas North's argument relies on a problematic criterion for comparing structure---based on the size of symmetry groups \parencites[]{Swanson}---our argument avoids this pitfall. We show that a system needs to satisfy strictly fewer physical conditions in order to be Hamiltonian. To be Lagrangian, systems must satisfy an additional physical condition, making Lagrangian mechanics a proper subset of Hamiltonian mechanics. Hamiltonian mechanics thereby captures a wider range of systems than Lagrangian mechanics, and this difference has interpretive consequences even within the hyper-regular domain. The construction at the heart of our argument provides an interpretation of classical mechanical systems in terms of their invariant phase space density. We show that the invariant density sits more naturally within the Hamiltonian framework, providing another reason to privilege it over Lagrangian mechanics.
% Our formalism also provides a novel interpretation of the principle of stationary action within Hamiltonian mechanics.

Section~\ref{privileging} provides an informal overview of our argument for privileging Hamiltonian mechanics. We introduce three physical conditions that define classical systems while distinguishing Hamiltonian from Lagrangian mechanics. Section~\ref{fundamentality} explains how our position differs from North's. Specifically, our argument does not require metaphysical commitments to fundamental structure or perfectly natural properties. Section~\ref{derivation} describes in detail the derivation of Hamiltonian mechanics for systems that satisfy two physical conditions: infinitesimal reducibility and deterministic/reversible evolution. We then show (Section~\ref{motion}) that Lagrangian systems must satisfy an additional physical condition, which we call motion equivalence. This completes our primary argument for privileging Hamiltonian mechanics.

Subsequent sections provide additional reasons to privilege Hamiltonian mechanics, along with illustrating further interpretive benefits of our approach. Section~\ref{density} explains how the invariant phase space density plays a special role in the Hamiltonian framework. The density allows us to compare and track the amount of material in different regions of phase space. In contrast, Lagrangian mechanics does not support these important properties of the density. Despite being coordinate-invariant, the density itself is coordinate-dependent. We therefore argue that some quantities are physically significant despite being coordinate-dependent. Finally, Section~\ref{Lagrangian} considers the status of Lagrangian mechanics within our framework. We provide a geometric interpretation of the action principle, which applies even outside the hyper-regular domain. We also consider and rebut objections stemming from Curiel's \parencites*[]{Curiel} privileging of Lagrangian mechanics. Along the way, we consider both the gauge redundancy of the conjugate momentum (Section~\ref{fundamental}) and of the Lagrangian (Section~\ref{gauge}).

% We define classical systems in terms of basic physical conditions that they satisfy. These conditions justify the appropriateness of certain mathematical structures.

% Understanding the relationship between Hamiltonian and Lagrangian mechanics depends partly on how one defines ``classical systems." Different definitions can result in different interpretive relationships. For instance, restricting classical systems to the hyper-regular domain precludes certain interpretive considerations. To avoid verbal disputes, it is essential to clarify what falls within the scope of ``classical systems." One way to delimit classical systems is to specify a set of equations that they satisfy---such as Hamilton's equations or the Euler--Lagrange equations. We will instead follow a different approach by delimiting classical systems based on the physical conditions that they satisfy. Section~\ref{infinitesimal} introduces the condition of \textit{infinitesimal reducibility}, meaning that the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (we cannot specify the whole state by specifying the states of its subsystems). In the widest sense, a classical system is any infinitesimally reducible system.
%a reducible system comprises subsystems whose states jointly determine the state of the whole system.

% Section~\ref{deterministic} introduces the joint condition of deterministic and reversible evolution. We will show that a classical system which satisfies this condition is Hamiltonian: its phase space is a symplectic manifold that evolves according to Hamiltonian evolution  (i.e. according to a symplectomorphism). Following a deterministic and reversible evolution is equivalent to being a conservative system, i.e. one that conserves energy. Hence, our discussion of Hamiltonian mechanics necessarily excludes dissipative systems, i.e. ones that do not conserve energy. Likewise, our derivation of Lagrangian mechanics in Section~\ref{motion} will also take place within the context of conservative systems. Consequently, we will not consider systems that satisfy inhomogenous Euler--Lagrange equations. We, of course, do not deny the sense in which the Lagrangian formulation applies to dissipative systems, but this is not what we will mean by ``Lagrangian mechanics" \parencites[]{Smith}{Curiel}.\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian.

@@

@@

Our argument has three main steps, summarized in Figure~\ref{diagram}. The left column lists three basic physical conditions: infinitesimal reducibility, determinism/reversibility, and motion equivalence. Each condition entails physically significant properties, described in the middle column. Each physical property corresponds to a formal mathematical framework depicted in the right column: including phase space, Hamiltonian mechanics, and Lagrangian mechanics. In this way, our framework develops a correspondence between physical conditions and their associated mathematical formalisms. We thereby gain a straightforward interpretation of the physical significance of these formalisms.
%Of the various notions of fundamentality, there is at least the following non-mysterious definition:

@@

First, a system is \textit{infinitesimally reducible} provided that the state of the whole system is equivalent to specifying the states of its parts (and vice versa). Infinitesimal reducibility distinguishes classical from quantum systems, where the latter are irreducible (we cannot specify the whole state by specifying the states of its subsystems). In the widest sense, a classical system is any infinitesimally reducible system. Colloquially, we can say that the whole state is equal to the sum of its parts, where we take these infinitesimal parts to be classical particles.

If a system is infinitesimally reducible, then we can define a distribution $\rho$ over particle states. Furthermore, since particle states are independent of a choice of units, this distribution must be invariant. We show in Section~\ref{infinitesimal} that to define this invariant distribution, we require a symplectic manifold, which serves as the phase space of classical mechanics. The physical significance of this symplectic manifold is thereby its connection with the physical condition of infinitesimal reducibility.

[To do: Clarify the scope of our treatment: at least in this paper, it seems to be restricted to $n$-particle classical mechanics. Does it also handle rigid bodies and classical continuum mechanics? Cite Mark Wilson 2013 on these issues]. 
% % I think I could spend more time clarifying reducibility, infinitesimal. Maybe define separately first?

Secondly, a system has \textit{deterministic and reversible evolution} provided that the present state of the system uniquely determines both its future and past states. This condition entails that the invariant distribution $\rho$ is in fact a conserved phase space density. We will show that a classical system which satisfies this condition is Hamiltonian: its phase space is a symplectic manifold that evolves according to Hamiltonian evolution  (i.e. according to a symplectomorphism). We thereby characterize Hamiltonian mechanics as the mathematical framework for infinitesimally reducible systems with deterministic and reversible evolution. Section~\ref{deterministic} justifies this correspondence.

Finally, some classical systems satisfy a further condition, which we call \textit{motion equivalence}. Motion equivalence holds when the system's trajectory in physical space determines---and is determined by---its evolution in phase space. In other words, knowledge of the trajectory in physical space (e.g. Euclidean three-space $\mathbb{R}^3$) suffices for knowledge of the trajectory in phase space, and vice versa. Physically, knowledge of the trajectory in physical space corresponds to knowledge of position and velocity, while knowledge of the trajectory in phase space requires knowing the system's momentum. Hence, motion equivalence entails a correspondence between velocity and momentum, which formally leads to the framework of Lagrangian mechanics in the hyper-regular domain (Section~\ref{motion}). As a result, Lagrangian mechanics requires that a further physical condition is satisfied, and hence it is less general than Hamiltonian mechanics. 
% % Would be ideal if we can provide some literature support for this notion of motion equivalence. E.g. connecting it with a discussion in Marsden and Abraham foundations of mechanics.

We cannot always determine the evolution of a system in phase space from its trajectory in physical space, since some systems fail to satisfy the condition of motion equivalence. In general, knowing a system's position and velocity does not suffice for knowing its momentum. Considering a photon as a classical particle, we cannot determine its momentum from its trajectory through space. Hence, knowledge of the physical trajectory of a photon is insufficient for knowledge of its phase space evolution. Different Lagrangians---differing even by more than a constant---can correspond to the same particle trajectories while not agreeing on the total energy, even up to a constant.\footnote{\textcites[1185--1186]{Barrett2} discusses this in the context of the failure of equivalence between the vector field formulations of Lagrangian and Hamiltonian mechanics. In contrast, the trajectories in phase space (encoded by the Hamiltonian vector field) uniquely define the Hamiltonian up to a constant.} Hence, the set of classical systems that satisfy motion equivalence is a proper subset of the classical systems that satisfy the conditions for Hamiltonian mechanics. For at least this reason, Hamiltonian mechanics is more fundamental than Lagrangian mechanics.
% [To do: verify that the scenario of two different Lagrangians---giving rise to the same trajectories in physical space with different energies---arises ONLY when the motion equivalence condition fails. See Curiel 2014, page 42 for his take on this kind of scenario.]

Of course, understanding the relationship between Hamiltonian and Lagrangian mechanics partly depends on how one defines ``classical systems." Different definitions can result in different interpretive relationships. On our account, Newtonian systems are the widest domain of classical systems, including any system that is infinitesimally reducible. Hamiltonian mechanics picks out the subset of Newtonian systems that follow deterministic and reversible evolution. Our treatment thereby sets aside well-known cases of non-deterministic classical systems.\footnote{For discussion of non-deterministic classical systems, see \parencites*[]{Earman,Baez,Norton}.} Additionally, we set aside dissipative systems, i.e. those that fail to conserve energy. Possessing a deterministic and reversible evolution is equivalent to being non-dissipative, i.e. conserving energy. Consequently, our derivation of Lagrangian mechanics in Section~\ref{motion} takes place within the context of conservative systems, so we do not consider systems that satisfy inhomogenous Euler--Lagrange equations. Although the Lagrangian formulation applies to dissipative systems \parencites[]{Smith,Curiel}, this is not what we will mean by ``Lagrangian systems."\footnote{Curiel \parencites*[311]{Curiel} acknowledges that the Hamiltonian framework can be extended to treat dissipative systems, although he seems to view it as a virtue of the Lagrangian formulation that it handles dissipative systems with fewer modifications.} Hence, our claim about the fundamentality of Hamiltonian mechanics is restricted to the non-dissipative regime. We will view non-dissipative systems as Newtonian, but neither Lagrangian nor Hamiltonian. Before presenting the derivational details of our argument in Section~\ref{derivation}, we clarify what we mean by ``fundamental."
%To avoid verbal disputes, it is essential to clarify what falls within the scope of ``classical systems." One way to delimit classical systems is to specify a set of equations that they satisfy---such as Hamilton's equations or the Euler--Lagrange equations. We will instead follow a different approach by delimiting classical systems based on the physical conditions that they satisfy.

 %[To do: on the topic of failures of determinism, cite Earman  \textit{a primer on determinism}, Baez singularities paper, or Norton's dome paper].

@@

% %Overall summary: A classical system is an infinitesimally reducible system. The state space of these systems can be modeled by a symplectic manifold (the physical assumption of infinitesimal reducibility corresponds to the mathematical structure of symplectic manifolds). Classical Hamiltonian systems satisfy the additional conditions of having deterministic and reversible evolutions. These two physical assumptions entail the existence of a symplectic morphism on the state space manifold, making it a contact manifold. Classical Lagrangian systems satisfy a further assumption: keeping track of the trajectories in physical space is equivalent to tracking the evolution of states in phase space.

\section{Fundamentality, \textit{sans} metaphysics}
\label{fundamentality}

% % done; slip in a citation to Dorr/Hawthorne 2014, to note the independent metaphysical challenges that the notion of naturalness faces, and likewise for a metaphysical notion of fundamentality.

@@

As indicated at the start of Section~\ref{privileging}, we use ``fundamentality" in a non-metaphysical fashion. On our usage, differences in fundamentality amount to differences in generality.\footnote{We also embrace other non-metaphysical uses of ``fundamental," tracking differences in epistemic or conceptual primacy. For instance, we can think of the canonical position and momentum observables as ``epistemically fundamental" because they suffice to completely specify a classical state, fixing its values for all other observables. \textcites[31, 200]{Ruetsche} refers to this as ``fundamental in the physicist's sense." Section~\ref{fundamental} discusses further senses in which the Hamiltonian formalism is more fundamental than Lagrangian mechanics.} Our notion of fundamentality leads to a correspondingly non-metaphysical notion of ``structure." Hamiltonian mechanics employs less structure than Lagrangian mechanics because the former requires fewer physical conditions. Compared with North's \parencites*[]{North} account of structure, our definition has two distinct advantages. First, it avoids metaphysical disputes surrounding perfectly natural or joint-carving properties, which empiricists and pragmatists decry. Secondly, it avoids a technical objection that Swanson and Halvorson \parencites*[]{Swanson} levy against North's more precise mathematical criterion for comparing structure.

Despite our different commitments, our approach has many methodological parallels to North's. We agree with North that it is fruitful to identify both invariants and ``the mathematical structure needed to formulate the theory in an invariant, frame-independent way" \parencites*[65]{North}. Additionally, we agree that it is worthwhile to identify what particular mathematical structure is required by a formulation \parencites[78]{North}. Indeed, we undertake precisely this sort of investigation in Section~\ref{derivation}. However, North advocates a metaphysical account of structure, about which we remain agnostic. According to North, ``structure comprises the objective, fundamental, intrinsic features, the ones that remain the same regardless of arbitrary or conventional choices and description" \parencites*[66]{North}. These notions of fundamentality and intrinsicality correspond to differences in joint-carving, where more fundamental or intrinsic features are more joint-carving (or in Lewis's \parencites*[]{Lewis1983} terminology, more natural).

Owing to our epistemic inabilities to access these putative joints, we do not endorse these metaphysical dimensions of structure or fundamentality.\footnote{For a detailed discussion of difficulties facing the notion of metaphysical naturalness, see \textcites[]{Dorr_Hawthorne}.} While we will argue that Hamiltonian mechanics is significant partly because it elucidates invariant distributions (Section~\ref{density}), this does not entail that additional mathematical structure is physically insignificant or surplus structure. Additional structure may correspond to an additional physically-significant assumption that applies to the system under consideration. Precisely this scenario occurs in the context of Lagrangian mechanics, which requires the additional condition of motion equivalence to hold. Whereas North views Lagrangian mechanics as containing ``excess, superfluous structure" \parencites*[75]{North}, we take this additional condition as unnecessary but still physically significant. Thus, while it is illuminating to identify the minimum mathematical structures necessary for a formulation, additional structure is not automatically rendered physically insignificant. %It corresponds to the physical assumption of motion equivalence, which applies to many classical systems. 
%This characterizes the relationship between Hamiltonian and Lagrangian formalizations of a classical system. 
% classical mechanics also deploys physically significant representations that are not intrinsic to mechanical systems [to do: mention some examples, e.g. in a footnote. The Lagrangian? Cartesian coordinates?].

@@

Additionally, whereas North \parencites*[76]{North} seeks to determine the ``real" fundamental state space structure of the theory---and correspondingly, of a classical world---we remain neutral on this underlying metaphysical question.\footnote{Indeed, Swanson and Halvorson \parencites*[]{Swanson} note that many physicists and philosophers would not interpret statespace structure as part of the theory's ontology, but as a formal device for encoding physical information or as a calculational tool.} Whether Hamiltonian phase space structure is metaphysically fundamental (in a classical world) does not matter for the questions we seek to answer. Regardless of its metaphysical status, Hamiltonian mechanics is physically more fundamental, and we will argue for its epistemic and conceptual primacy as well. Indeed, our argument does not privilege any one of the at least three different mathematical ways of formalizing Hamiltonian mechanics [to do: mention these different Hamiltonian formalisms and provide references]. Each of these frameworks has different pragmatic costs and benefits. Hence, it is plausible that there may be no single, uniquely necessary mathematical structure for mechanics, but rather a family of related structures. In other words, there may be no canonical minimal structure for formalizing the relevant physical models and physical assumptions.\footnote{Tulczyjew's reformulation of Hamiltonian mechanics supplies a further reason for viewing symplectic geometry---and hence Hamiltonian mechanics---as conceptually fundamental. According to Teh and Tsementzis, this reformulation shows that  ``the very concept of a classical mechanical system (and not merely the phase space of that system) should be described in terms of symplectic geometry" \parencites*[46]{Teh}.}  Additionally, each of these formalisms references some ``unphysical" coordinates or representational means, whose physical significance is borne out by their fruitfulness in modeling. We return to the physical significance of certain coordinate choices in Section~\ref{fundamental}.
%The mathematical structure that you ``need" might not be unique. Neither one of these three formalizations of Hamiltonian mechanics is necessarily more ``fundamental" or joint-carving than the others. Nevertheless, there are compelling physical and pragmatic motivations for preferring the extended phase space formulation of Hamiltonian mechanics. In this formulation, we gain an invariant Hamiltonian that is the conjugate variable of the proper time, while the normal Hamiltonian (which is not coordinate-invariant) still plays the role of generating time-translations.

North also introduces a technical notion of comparative structure, based on the symmetry group of statespace. According to this criterion, statespace A has more structure than statespace B if the symmetry group of A is a proper subgroup of B's symmetry group \parencites[87-88]{North}.\footnote{This criterion follows from the most charitable reconstruction of North's argument, given by \textcites[]{Swanson}. North sometimes speaks in terms of comparing the dimensions of symmetry groups, which they note leads to counterexamples.} A symmetry group that is a proper subgroup entails that strictly fewer transformations preserve the space's structure. This indicates that statespace A has more mathematical structure to be preserved. These symmetry groups are defined in terms of the coordinate transformations that preserve the space's geometric structure, amounting to re-coordinatizations (redescriptions) of that structure. For Lagrangian systems, North takes the relevant symmetry group to be that of the point transformations, and for Hamiltonian systems, she assumes that canonical transformations play this role. Since point transformations are a proper subgroup of canonical transformations, North's criterion entails that the statespace of Lagrangian mechanics has more structure than the statespace of Hamiltonian mechanics---in virtue of the former having a smaller symmetry group.
